---
title: "Profiling insular tetrapods vulnerable to biological invasions"
author: "Clara Marino"
date: "20/04/2021"
output: html_document
---


#### --------------------------------------------------------------
## Step 0: setup ##
#### --------------------------------------------------------------

Clean workspace & set up required packages

```{r}
# clean workspace
rm(list=ls())

# packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, tibble, tidyr, rlist,
               ape, clue, cluster, geometry, gtools, 
               ggplot2, ggthemes)
# dplyr, tidyr, tibble, miceadds, rlist: used for data manipulation 
# ape, clue, cluster, geometry, gtools: for FD metrics computation
# ggplot2, ggthemes: data visualization 

```

#### --------------------------------------------------------------
## Step 1: load data & functions ##
#### --------------------------------------------------------------

Load data

```{r}
# load trait data for each taxa
data_ft_amph <- readRDS("Data/1_data_ft_amph")
data_ft_birds <- readRDS("Data/2_data_ft_birds")
data_ft_mam <- readRDS("Data/3_data_ft_mam")
data_ft_rept <- readRDS("Data/4_data_ft_rept")

# load species groups (IAS/Other-threat, Threatened/Non-threatened)
df_insul <- readRDS("Data/5_df_insular_assembl")

```

Source functions for species to FE calculation & randomization

```{r}
# Source Villeger (2017) functions
source("FD_functions/species_to_FE.R")
# Source function for randomizing assemblages (for null models)
source("FD_functions/rand_vect_geom.R")

```


#### --------------------------------------------------------------
## Step 2: Null models on trait distribution among species ##
#### --------------------------------------------------------------

```{r}
# ------------------   Prepare assemblages   -----------------------

# Create a list with 4 dataframes with traits (one for each class)
list_data_ft <- list(Amphibia = data_ft_amph, Aves = data_ft_birds,
                     Mammalia = data_ft_mam, Reptilia = data_ft_rept)

# Create a list with 4 assemblages (one for each class)
list_assembl <- lapply(list_data_ft, function(x){
  data_nw <- df_insul %>% 
    filter(binomial %in% x$binomial) %>%
    distinct(binomial, group)
  tb <- as.data.frame(table(data_nw$group, as.character(data_nw$binomial)))
  mat_assemb <- spread(tb, Var2, Freq)
  rownames(mat_assemb) <- mat_assemb$Var1
  mat_assemb$Var1 <- NULL
  mat_assemb <- as.matrix(mat_assemb)
  return(mat_assemb)})

# Remove species names to keep only traits in data_ft
list_data_ft <- lapply(list_data_ft, function(x){
  row.names(x) <- x$binomial
  x <- x %>% select(-binomial)})

# --------------------  Group species to FEs  ----------------------

# Grouping species into FEs
list_sp_to_FE <- lapply(list_data_ft, species_to_FE)

# trait values of the FE
list_FE_traits <- lapply(list_sp_to_FE, function(x){x$FE_traits})

# Matrix of FE in assemblages
list_assembl_FE <- mapply(x=list_assembl, y=list_sp_to_FE, function(x,y){
  # Empty matrix
  assembl_FE<-matrix(0, nrow(x), nrow(y$FE_sp_01), 
                     dimnames=list(row.names(x), row.names(y$FE_sp_01))) 
  # loop on FEs
  for (k in row.names(y$FE_sp_01)){
    sp_k<-names(which(y$FE_sp_01[k,]==1))   
    if( length(sp_k)==1 ) {assembl_FE[,k]<-x[,sp_k] } # if only one species in FE k
    if (length(sp_k)>1 ) {assembl_FE[,k]<-apply(x[,sp_k],1,sum)  } # if more than 1 species in FE k
  }# end of k
  return(assembl_FE)
})


# -----------------  Simulate n possible assemblages  -------------------

# number of iterations
n = 99 # takes 20 min for n=9999

Sys.time()
list_simulated_FE <- lapply(list_assembl_FE, function(x){
  
  # initialisation with the first simulated assemblage
  random_mat_0 <- as.data.frame(x)
  colnames(random_mat_0) <- sample(colnames(random_mat_0))
  random_mat_0 <- as.matrix (random_mat_0)
  N_sp <- rowSums(random_mat_0)
  N_FE <- rowSums(random_mat_0!=0)
  
  for (k in 1:nrow(random_mat_0)){
    random_mat_0[k,][which(random_mat_0[k,]!=0)] <- rand_vect_geom(N_FE[k], N_sp[k])
  }
  random_mat_0 <- random_mat_0[,order(colnames(random_mat_0))]
  
  # randomization & row bind of initial assemblage + n-1 simulated assemblages
  for (i in 1:(n-1)){
    random_mat <- as.data.frame(x)
    colnames(random_mat) <- sample(colnames(random_mat))
    random_mat <- as.matrix (random_mat)
    N_sp <- rowSums(random_mat)
    N_FE <- rowSums(random_mat!=0)
    
    for (k in 1:nrow(random_mat)){
      random_mat[k,][which(random_mat[k,]!=0)] <- rand_vect_geom(N_FE[k], N_sp[k])
    }
    random_mat <- random_mat[,order(colnames(random_mat))]
    
    random_mat_0 <- rbind(random_mat_0, random_mat)
  }
  
  return(random_mat_0)}) 
Sys.time()

list_simulated_FE <- lapply(list_simulated_FE, as.data.frame)

# -----------------  Count observed modalities -------------------

# Create a list with 4 df traits (one for each class)
list_data_ft <- list(Amphibia = data_ft_amph, Aves = data_ft_birds,
                     Mammalia = data_ft_mam, Reptilia = data_ft_rept)


observed_modalities <- lapply(list_data_ft, function(x){
  
  # add a column for threat_status group (0_0, 0_1, 1_0, 1_1)
  df <- left_join(x, df_insul %>% select(binomial, group),
                  by="binomial")
  
  # initialise table for the first trait (first col after binomial)
  to_bind <- as.data.frame(table(df$group, df[,2])) %>%
    mutate(Trait = names(df)[2])
  # table on each trait (all columns except binomial & threatened_ias)
  for (i in 3:(length(names(df))-1)) {
    to_bind2 <- as.data.frame(table(df$group, df[,i])) %>%
      mutate(Trait = names(df)[i])
    to_bind <- bind_rows(to_bind, to_bind2)
  }
  
  # wanted columns: Group, Trait, Modality, Obs_value
  obs_val <- to_bind %>%
    dplyr::rename(Group = Var1,
           Modality = Var2,
           Obs_value = Freq)
  return(obs_val)
  })


# -----------------  Save outputs -------------------

# simulated FE
saveRDS(list_simulated_FE, paste0('Trait_FE_tetra_sim_', n))
# species x FE correspondances 
saveRDS(list_sp_to_FE, 'sp_to_FE_tetra')
# observed trait modalities
saveRDS(observed_modalities, 'Trait_tetra_obs')

```


#### --------------------------------------------------------------
## Step 3: Compare observed and simulated trait values ##
#### --------------------------------------------------------------

Compute ES and p-values for IAS-T and IAS-NT assemblages

```{r}

# Clean workspace
rm(list=ls())

# source function for p calculation
source("FD_functions/calc_p_val.R")

# Load outputs from Step 2
n=99 # p-calc works only for large n (n>500)
sim_trait_FE <- readRDS(paste0('Trait_FE_tetra_sim_', n))
sp_to_FE <- readRDS("sp_to_FE_tetra")
obs_trait <- readRDS("Trait_tetra_obs")

# --------------- Shape FE_traits & sim_trait_FE ------------------

# order rownames
FE_traits <- lapply(sp_to_FE, function(x){
  x$FE_traits[order(rownames(x$FE_traits)),]}) 

# transpose matrix
t_rand_trait <- lapply(sim_trait_FE, function(x){t(x)})

# --------------- Calculate simulated means & sd ------------------

# Compute null models for each assemblage type separately

# For IAS-NT species (Group "IAS_NT")
IAS0_traits <- mapply(x = FE_traits, y = t_rand_trait, function(x,y){
  cbind(x %>%
          # pass ordered factors to simple factors
          mutate(Habitat = factor(Habitat, ordered = F),
                 Body.mass = factor(Body.mass, ordered = F)),
        as.data.frame(y) %>% 
          select(contains('IAS_NT'))) # select IAS-NT assemblages only 
})
ag_IAS0 <- lapply(IAS0_traits, function(x){
  IAS0_traits_g <- x %>% 
    pivot_longer(c(Habitat:For.niche), names_to = "Trait", values_to = "Modality")
  ag_IAS0 <- aggregate(.~Trait+Modality, data = IAS0_traits_g, FUN=sum)
  return(ag_IAS0)
})
null_mod_IAS0 <- lapply(ag_IAS0, function(x){
  null_mod_IAS0 <- cbind(x[,1:2])
  null_mod_IAS0$meanNM <- apply(x[,c(3:(n+2))], 1, mean)
  null_mod_IAS0$sdNM <- apply(x[,c(3:(n+2))], 1, sd)
  null_mod_IAS0$status <- c('IAS_NT')
  return(null_mod_IAS0)
})

# For IAS-T species (Group "IAS_T")

IAS1_traits <- mapply(x = FE_traits, y = t_rand_trait, function(x,y){
  cbind(x %>%
          # pass ordered factors to simple factors
          mutate(Habitat = factor(Habitat, ordered = F),
                 Body.mass = factor(Body.mass, ordered = F)),
        as.data.frame(y) %>% 
          select(contains('IAS_T'))) # select IAS-T assemblages only 
})
ag_IAS1 <- lapply(IAS1_traits, function(x){
  IAS1_traits_g <- x %>% 
    pivot_longer(c(Habitat:For.niche), names_to = "Trait", values_to = "Modality")
  ag_IAS1 <- aggregate(.~Trait+Modality, data = IAS1_traits_g, FUN=sum)
  return(ag_IAS1)
})
null_mod_IAS1 <- lapply(ag_IAS1, function(x){
  null_mod_IAS1 <- cbind(x[,1:2])
  null_mod_IAS1$meanNM <- apply(x[,c(3:(n+2))], 1, mean)
  null_mod_IAS1$sdNM <- apply(x[,c(3:(n+2))], 1, sd)
  null_mod_IAS1$status <- c('IAS_T') 
  return(null_mod_IAS1)
})


# Combine results for IAS-T & IAS-NT assemblages in one df 
# -->  not possible to do rbind in mapply function
# need to use for loop across all list elements
df_nm <- FE_traits
for (i in 1:4){
  df_nm[[i]] <- bind_rows(null_mod_IAS0[[i]], null_mod_IAS1[[i]]) %>%
    mutate(key = paste(Trait, Modality, status, sep='_'))
}

ag_all <- ag_IAS0
for (i in 1:4){
  ag_IAS0[[i]]$status <- c('IAS_NT')
  ag_IAS1[[i]]$status <- c('IAS_T')
  colnames(ag_IAS0[[i]]) <- colnames(ag_IAS1[[i]])
  ag_all[[i]] <- bind_rows(ag_IAS0[[i]], ag_IAS1[[i]]) %>%
    mutate(key = paste(Trait, Modality, status, sep='_'))
}

# ----------------  Observed vs. simulated values  -------------------

# Shape observed values
obs_values <- lapply(obs_trait, function(x){
  x %>%
    mutate(key = paste(Trait, Modality, Group, sep='_')) %>%
    select(key, Obs_value)
})

df_p_values <- df_nm
traits_nm_final <- df_nm

for (i in 1:4){
  
  # Calculate p.values
  
  df_p_values[[i]] <- inner_join(obs_values[[i]], ag_all[[i]], by='key') %>% 
    select(-Trait, -Modality, -status)
  
  df_p_values[[i]]$p.val <- 0
  df_p_values[[i]]$b <- 0
  
  for (r in 1:nrow(df_p_values[[i]])){
    
    sim <- as.numeric(df_p_values[[i]][r,] %>%
                        select(-c("key","Obs_value","p.val","b")))
    obs <- as.numeric(df_p_values[[i]]$Obs_value[r])
    
    b = sum(obs<sim)+sum(obs==sim)
    N = length(sim)
    
    if (b<6||b>(n-5)) {
      fit <- density(sim)
      all_p <- vector()
      for(w in 1:1000){
        x.new <- rnorm(N, sim, fit$bw)
        all_p[w] <- calc_pval_trait(x.new, obs)
      }
      p = median(all_p)
    } else {
      p = (b+1)/(N+1)
    }
    
    # store p-val and b to check
    df_p_values[[i]]$p.val[r] <- p
    df_p_values[[i]]$b[r] <- b
  }
  
  # join all df to have one clean output
  # calculate ES
  traits_nm_final[[i]] <- df_p_values[[i]] %>%
    select(key, Obs_value, b, p.val) %>%
    mutate(ES = qnorm(1-p.val)) 
}

# ----------------  Save final output -------------------

saveRDS(traits_nm_final, "Trait_tetra_nm_es_pval")

```


Compute ES and p-val for Other-threat-T assemblage

```{r}
# Clean workspace
rm(list=ls())

# source function for p calculation
source("FD_functions/calc_p_val.R")

# Load outputs from Step 2
n=99 # p-calc works only for large n (n>500)
sim_trait_FE <- readRDS(paste0('Trait_FE_tetra_sim_', n))
sp_to_FE <- readRDS("sp_to_FE_tetra")
obs_trait <- readRDS("Trait_tetra_obs")

# --------------- Shape FE_traits & sim_trait_FE ------------------

# order rownames
FE_traits <- lapply(sp_to_FE, function(x){
  x$FE_traits[order(rownames(x$FE_traits)),]}) 

# transpose matrix
t_rand_trait <- lapply(sim_trait_FE, function(x){t(x)})

# --------------- Calculate simulated means & sd ------------------

# Compute null models for each assemblage type separately

# For other threats NT species (Group "other_NT")
other0_traits <- mapply(x = FE_traits, y = t_rand_trait, function(x,y){
  cbind(x %>%
          # pass ordered factors to simple factors
          mutate(Habitat = factor(Habitat, ordered = F),
                 Body.mass = factor(Body.mass, ordered = F)),
        as.data.frame(y) %>% 
          select(contains('other_NT'))) # select other-NT assemblages only 
})
ag_other0 <- lapply(other0_traits, function(x){
  other0_traits_g <- x %>% 
    pivot_longer(c(Habitat:For.niche), names_to = "Trait", values_to = "Modality")
  ag_other0 <- aggregate(.~Trait+Modality, data = other0_traits_g, FUN=sum)
  return(ag_other0)
})
null_mod_other0 <- lapply(ag_other0, function(x){
  null_mod_other0 <- cbind(x[,1:2])
  null_mod_other0$meanNM <- apply(x[,c(3:(n+2))], 1, mean)
  null_mod_other0$sdNM <- apply(x[,c(3:(n+2))], 1, sd)
  null_mod_other0$status <- c('other_NT')
  return(null_mod_other0)
})

# For other threats T species (Group "other_T")

other1_traits <- mapply(x = FE_traits, y = t_rand_trait, function(x,y){
  cbind(x %>%
          # pass ordered factors to simple factors
          mutate(Habitat = factor(Habitat, ordered = F),
                 Body.mass = factor(Body.mass, ordered = F)),
        as.data.frame(y) %>% 
          select(contains('other_T'))) # select other-T assemblages only 
})
ag_other1 <- lapply(other1_traits, function(x){
  other1_traits_g <- x %>% 
    pivot_longer(c(Habitat:For.niche), names_to = "Trait", values_to = "Modality")
  ag_other1 <- aggregate(.~Trait+Modality, data = other1_traits_g, FUN=sum)
  return(ag_other1)
})
null_mod_other1 <- lapply(ag_other1, function(x){
  null_mod_other1 <- cbind(x[,1:2])
  null_mod_other1$meanNM <- apply(x[,c(3:(n+2))], 1, mean)
  null_mod_other1$sdNM <- apply(x[,c(3:(n+2))], 1, sd)
  null_mod_other1$status <- c('other_T')
  return(null_mod_other1)
})

# Combine results for 1_0 & 0_0 assemblages in one df 
# -->  not possible to do rbind in mapply function
# need to use for loop across all list elements
df_nm <- FE_traits
for (i in 1:4){
  df_nm[[i]] <- bind_rows(null_mod_other0[[i]], null_mod_other1[[i]]) %>%
    mutate(key = paste(Trait, Modality, status, sep='_'))
}

ag_all <- ag_other0
for (i in 1:4){
  ag_other0[[i]]$status <- c('other_NT')
  ag_other1[[i]]$status <- c('other_T')
  colnames(ag_other0[[i]]) <- colnames(ag_other1[[i]])
  ag_all[[i]] <- bind_rows(ag_other0[[i]], ag_other1[[i]]) %>%
    mutate(key = paste(Trait, Modality, status, sep='_'))
}

# ----------------  Observed vs. simulated values  -------------------

# Shape observed values
obs_values <- lapply(obs_trait, function(x){
  x %>%
    mutate(key = paste(Trait, Modality, Group, sep='_')) %>%
    select(key, Obs_value)
})

df_p_values <- df_nm
traits_nm_final <- df_nm

for (i in 1:4){
  
  # Calculate p.values
  
  df_p_values[[i]] <- inner_join(obs_values[[i]], ag_all[[i]], by='key') %>% 
    select(-Trait, -Modality, -status)
  
  df_p_values[[i]]$p.val <- 0
  df_p_values[[i]]$b <- 0
  
  for (r in 1:nrow(df_p_values[[i]])){
    
    sim <- as.numeric(df_p_values[[i]][r,] %>%
                        select(-c("key","Obs_value","p.val","b")))
    obs <- as.numeric(df_p_values[[i]]$Obs_value[r])
    
    b = sum(obs<sim)+sum(obs==sim)
    N = length(sim)
    
    if (b<6||b>(n-5)) {
      fit <- density(sim)
      all_p <- vector()
      for(w in 1:1000){
        x.new <- rnorm(N, sim, fit$bw)
        all_p[w] <- calc_pval_trait(x.new, obs)
      }
      p = median(all_p)
    } else {
      p = (b+1)/(N+1)
    }
    
    # store p-val and b to check
    df_p_values[[i]]$p.val[r] <- p
    df_p_values[[i]]$b[r] <- b

  }
  
  # join all df to have one clean output
  # calculate ES
  traits_nm_final[[i]] <- df_p_values[[i]] %>%
    select(key, Obs_value, b, p.val) %>%
    mutate(ES = qnorm(1-p.val)) 

}
# ----------------  Save final output -------------------

saveRDS(traits_nm_final, "Trait_tetra_nm_es_pval_other_threat")


```

#### --------------------------------------------------------------
## Step 4: Plot outputs ##
#### --------------------------------------------------------------

Figure representing dominant and overrepresented modalities of IAS-T tetrapods have been realized manually using Inskape and results from "traits_nm_final" table.

Figure comparing traits of IAS-T vs IAS-NT species:

```{r}

es <- readRDS("Trait_tetra_nm_es_pval")

# ------------ Traits of IAS-T vs IAS-NT species ----------------

es_clean <- lapply(es, function(x){
  x %>% 
    # change multi_anim to avoid"_" and separate "key" column
    mutate(key=recode(key, 
                      "Main.diet_Multi_anim_IAS_NT"="Main.diet_MultiAnim_IAS_NT",
                      "Main.diet_Multi_anim_IAS_T"="Main.diet_MultiAnim_IAS_T")) %>%
    # separate key colum to have Group, Trait & Modality var
    separate(key, c("Trait","Modality","Group"), "_", extra = "merge") %>%
    # create a col for plotting deviation beatween mean & observed value
    mutate_if(is.character, as.factor)%>% 
    select(Trait, Modality, Group, ES, p.val) %>%
    pivot_wider(names_from = Group, values_from = c(ES, p.val))
})

# Plot all ES (signif & non signif)
# All classes together

es_all <- bind_rows(es_clean, .id = "Class")

# select only signif ES
es_signif_all <- es_all %>% 
  filter(p.val_IAS_NT > 0.975 | p.val_IAS_T > 0.975 |
           p.val_IAS_T < 0.025 | p.val_IAS_NT < 0.025 )

ggplot(es_signif_all, aes(x=ES_IAS_T, y=ES_IAS_NT, group=Trait)) +
  geom_point(aes(color=Trait)) + 
  scale_color_manual(values=c("chartreuse3","steelblue2","darkorange",
                              "darkslategrey","slateblue3","red3"))+
  geom_text(aes(label=Modality),hjust=0, vjust=0)+
  xlim(-4, 4) + ylim(-4,4) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0) +
  theme_classic()+
  facet_wrap(~Class)

```

Figure comparing traits of IAS-T and Other-threat-T species:

```{r}

ias <- readRDS("Trait_tetra_nm_es_pval")

other <- readRDS("Trait_tetra_nm_es_pval_other_threat")

ias_clean <- lapply(ias, function(x){
  x %>% 
    # change multi_anim to avoid"_" and separate "key" column
    mutate(key=recode(key, 
                      "Main.diet_Multi_anim_IAS_NT"="Main.diet_MultiAnim_IAS_NT",
                      "Main.diet_Multi_anim_IAS_T"="Main.diet_MultiAnim_IAS_T")) %>%
    # separate key colum to have Group, Trait & Modality var
    separate(key, c("Trait","Modality","Group"), "_", extra = "merge") %>%
    # create a col for plotting deviation between mean & observed value
    mutate_if(is.character, as.factor)%>% 
    select(Trait, Modality, Group, ES, p.val)
})

other_clean <- lapply(other, function(x){
  x %>% 
    # change multi_anim to avoid"_" and separate "key" column
    mutate(key=recode(key, 
                      "Main.diet_Multi_anim_other_NT"="Main.diet_MultiAnim_other_NT",
                      "Main.diet_Multi_anim_other_T"="Main.diet_MultiAnim_other_T")) %>%
    # separate key colum to have Group, Trait & Modality var
    separate(key, c("Trait","Modality","Group"), "_", extra = "merge") %>%
    # create a col for plotting deviation beatween mean & observed value
    mutate_if(is.character, as.factor)%>% 
    select(Trait, Modality, Group, ES, p.val)
})

all_clean <- ias_clean
for (i in 1:4){
  all_clean[[i]] <- bind_rows(ias_clean[[i]], other_clean[[i]])
}

es_all <- lapply(all_clean, function(x){
  x %>%
    pivot_wider(names_from = Group, values_from = c(ES, p.val))
})

es_all <- bind_rows(es_all, .id = "Class")

# select only signif SES
es_signif_all <- es_all %>% 
  filter(p.val_IAS_T > 0.975 | p.val_other_T > 0.975 |
           p.val_IAS_T < 0.025 | p.val_other_T < 0.025 )

ggplot(es_signif_all, aes(x=ES_IAS_T, y=ES_other_T, group=Trait)) +
  geom_point(aes(color=Trait)) + 
  scale_color_manual(values=c("chartreuse3","steelblue2","darkorange",
                              "darkslategrey","slateblue3","red3"))+ 
  geom_text(aes(label=Modality),hjust=0, vjust=0) +
  xlim(-4, 4) + ylim(-4,4) +
  geom_hline(yintercept=0) + geom_vline(xintercept=0) +
  stat_function(fun=function(x){x})+
  stat_function(fun=function(x){3*x}, color="red")+
  stat_function(fun=function(x){x/3}, color="red")+
  theme_classic()+
  facet_wrap(~Class)


```

